Thread Pool in GeckoDB/BOLSTER - Implementing a Thread Pool for a high performance Database System \\

### 2018-06-21 Presentations/Paper/paper_thread_pool.tex
Fix typos in paper

\title{Thread Pool in GeckoDB/BOLSTER - Integrating a Thread Pool into a high performance Database System \\
}

@@

\begin{abstract}
Processing tasks in parallel is used in nearly all applications to keep up with the requirements on modern software systems. However, the implementation of parallel processing can differ strongly depending on the desired use cases and reach from spawning dozens of threads to reuse threads in thread pools. In this paper, we show an implementation of a thread pool to process independent tasks in parallel including the waiting for a set of processed tasks. We also compare the thread pool implementation against a baseline, which spawns new threads for parallel data processing. The evaluation of the thread pool shows, that designing and implementing a thread pool is a balancing act between generalization of tasks and the processing performance. Additionally, we show that the task and thread pool configuration depending on the use case has a high impact on the thread pool performance.
\end{abstract}

@@

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. In this paper, we focus on multi threading by implementing a thread pool for the graph database system GeckoDB. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, many drawbacks arise from this implementation. 
First of all, the creation of threads comes along with overhead like stack initialisation and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is 	time-consuming.

To overcome these drawbacks, we integrate an optimised thread pool in BOLSTER. Along with the implementation, we measure the performance of the primitives to determine the thread pool overhead. Additionally, we measure metrics like \emph{waiting and busy time} of threads to evaluate correct thread pool sizes for the considered use cases.  In this work we make the following contributions:
\begin{itemize}
	\item We describe our design and implementation of the thread pool 
	\item We evaluate the possibility to wait for a group of task in the calling thread
	\item We compare our thread pool against the existing implementation in BOLSTER
\end{itemize}
We organized the rest of the paper as follows. In Section 2, we give preliminaries about the considered task configuration and about thread safe access of memory. In Section 3, we show our design and implementation of the thread pool and examine our experimental environment in Section 4. In Section 5, we describe the results of our performance evaluation. In Section 6, we name related work and state our conclusion and future work in Section 7.

@@

Each task can be enqueued with a priority. The priority zero is the highest and the task will be processed by the next free thread. The higher the priority of the task is, the further it is placed behind in the queue. Additionally, we only consider not preemtable tasks. Once a task is assigned to a thread, the thread will finish the operation of the task before getting a new one.

\subsection{Synchronizing Memory Access from Threads}
Parallelism with multi threading works great as long as each thread works on a separate memory area. During task scheduling, the scheduler has to know the state each thread has. This can be solved with signal handling or by writing the current thread state into main memory. In this work, we implement the second approach to avoid having another thread that only schedules tasks to other threads.

Thread safe access to memory can be achieved by different approaches. Firstly, a mutex thread can manage the access for multiple other threads. Secondly, atomic operations ensure the safe access from different threads to the same memory. We use both approaches for different parts of the thread pool. Since using a mutex thread can decrease access performance, we decided to execute atomic operations on memory containing the thread state informations. For example, the function \emph{atomic\_compare\_exchange\_strong} performs a compare and, if the result is true, an exchange of the memory in one atomic operation. We use a mutex thread to synchronize the enqueueing of tasks into the priority queue.

\section{Implementation}
In this section, we describe in detail the design and implementation of our thread pool. We show how tasks are enqueued regarding their priority and how the waiting for tasks is implemented. Additionally, we state our thread metrics and how they are integrated in the thread pool design.

@@

@@

Each assigned task contains a function pointer to a routine and a pointer to data. We implemented the task queue as a generic priority queue that uses a mutex thread to enqueue tasks threadsafe. Every time a new task is enqueued to the task queue, two scenarios can occur. If the new task has a priority, it is placed in the right slot of the task queue regarding its priority and the subsequent tasks are relocated. If the  new task has no priority, it is appended to the end of the queue. Furthermore, tasks can refer to a handle, which is used by the thread pool to wait for a specific amount of tasks.
\begin{table*}[htbp]
	\caption{Functions provided by the thread pool to process tasks}
	\begin{center}
		\begin{tabular}{ c c }
			\hline
			\textbf{Function Name}&\textbf{Explanation}\\
			\hline
			thread\_pool\_enqueue\_task & Enqueue task with optional handle \\
			thread\_pool\_enqueue\_tasks & Enqueue tasks with optional handle \\
			thread\_pool\_wait\_for\_task & Waits until the tasks referenced by the handle are completed. \\
			thread\_pool\_enqueue\_tasks\_wait & Enqueue tasks and wait until they are finished. The main thread also participates in task execution\\ 
			thread\_pool\_wait\_for\_all & Wait until all enqueued tasks are finished. The main thread also participates in task execution \\ \hline
		\end{tabular}
		\label{tab1}
	\end{center}
\end{table*}
\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

In the next step, the tasks will be enqueued into the priority queue of the thread pool. Before enqueueing, the function \emph{thread\_pool\_enqueue\_tasks\_wait} creates a handle and links all tasks to it. After enqueueing the tasks to the priority queue, the waiting threads in the thread pool pop the enqueued tasks from the priority queue and execute the routine of the task. This happens asynchronously to the calling function and therefore is not considered as an own step of the process. The calling threads gets one of the passed tasks and processes it. After finishing the task, the calling thread waits until all other tasks referring to the created handle are finished and then returns to the calling function. In the next section, we show the waiting for tasks with handles in detail.

@@

To solve this problem, we implement the waiting for a set of tasks using a slotmap. In Figure 3, we show the concept of our slotmap. During the enqueueing process of tasks, a set of tasks are referred to a slot in the slotmap. Each slot contains a number of open tasks to be finished and a generation counter. We name the number of open task \emph{openTaskCount} and the generation counter \emph{genCount}. Every time a set of tasks is enqueued, a slot with $openTaskCount = 0$ is selected and the \emph{genCount} increases. Every time a thread finishes a task, the $openTaskCount$ of the referred slot is decreased. As mentioned before, multiple tasks can refer to a single handle. This handle also refers to a slot and has additionally assigned a generation value. The waiting algorithm loops over the following conditions before returning to the calling function: It checks if the number of open tasks is zero or if the generation counter of this slot has changed. If one of these conditions is true, the waiting algorithm returns to the calling function.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different thread may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location another threads writes into the data, which results to inconsistency. In Section 2, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates. \\
 This approach enables efficient dynamic and lazy resizing of the thread pool. When the number of threads is reduced, the now redundant threads are marked to finish after completing their current task. If later on the number is increased again, still running but marked threads can be enabled again ensuring that few threads have to be created and that the number of running threads will not temporarily exceed the requested size. \\
 After presenting the enqueueing, processing, and waiting for tasks, we show the second component of our thread pool, the performance monitoring.

\subsection{Performance monitoring}
The performance monitoring component of the thread pool system is responsible for collecting data about threads, tasks, and the whole thread pool system. We will use the information collected in this component for analysis and performance optimization in later stages. As mentioned before, the performance monitoring is an optional component, since evaluating the following statistics can reduce the execution performance of our thread pool. We divide the collected statistics into the sections \emph{task statistics}, \emph{thread statistics}, and \emph{thread pool statistics}.

@@

How long a task waits in the queue before being executed is a good metric to adjust task priority or the thread pool size. For example, if an enqueued task with priority 0 still has a large difference between \emph{enqueue\_time} and \emph{execution\_time}, the thread pool size may be too small. Otherwise, if a large number of tasks without high priority have small differences between \emph{enqueue\_time} and \emph{execution\_time}, this may indicate that the thread pool size is too large. A better indicator of not optimized thread pool sizes are statistics per thread.

@@

The \emph{busy\_time} of a thread strongly depends on the executed tasks and therefor is a metric that gives less information about the thread pool itself. In contrast, the \emph{waiting\_time} is a good indicator to reveal less used threads in the thread pool to determine an optimal amount of threads for the current use case. Furthermore, the \emph{task\_count} of thread compared to the number of tasks processed by the thread pool indicates the degree of usage of the thread. To give further possibilities for optimizing the thread pool, we collect statistics for each thread pool instance.

>>>>

\title{Thread Pool in GeckoDB/BOLSTER - Implementing a Thread Pool for a high performance Database System \\
}

@@

\begin{abstract}
Processing tasks in parallel is used in nearly all applications to keep up with the requirements of modern software systems. However, the implementation of parallel processing can reach from spawning dozens of threads to reuse threads in thread pools. In this paper, we show an implementation of a thread pool to process independent tasks in parallel including the waiting for a set of processed tasks. We also compare the thread pool implementation against a baseline. The evaluation of the thread pool shows, that designing and implementing a thread pool is a balancing act between generalization of tasks and the processing performance. Additionally, we show that the task and thread pool configuration depending on the use case has a high impact on the thread pool performance.
\end{abstract}

@@

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. In this paper, we focus on multi threading by implementing a thread pool for the graph database system GeckoDB. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, a couple of drawbacks arise from this implementation. 
First of all, the creation of threads comes along with overhead like stack initialisation and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is 	time-consuming.

To overcome these drawbacks, we integrate an optimised thread pool in BOLSTER. Along with the implementation, we measure the performance of the primitives to determine the thread pool overhead. Additionally, we measure metrics like \emph{waiting and busy time} of threads to evaluate correct thread pool sizes for the considered use cases. In this work we make the following contributions:
\begin{itemize}
	\item We describe our design and implementation of the thread pool 
	\item We evaluate the possibility to wait for a group of task in the calling thread
	\item We compare our thread pool against the existing implementation in BOLSTER
	\item We measure and evaluate statistics for threads, tasks and the thread pool
\end{itemize}
We organized the rest of the paper as follows. In Section 2, we give preliminaries about the considered task configuration and about thread safe access of memory. In Section 3, we show our design and implementation of the thread pool and examine our experimental environment in Section 4. In Section 5, we describe the results of our performance evaluation. In Section 6, we name related work and state our conclusion and future work in Section 7.

@@

Each task can be enqueued with a priority. The priority zero is the highest and the task will be processed by the next free thread. The higher the priority of a task is, the further it is placed behind in the queue. Additionally, we only consider not preemtable tasks. Once a task is assigned to a thread, the thread will finish the operation of the task before getting a new one.

\subsection{Synchronizing Memory Access from Threads}
Parallelism with multi threading works great as long as each thread works on a separate memory location. During task scheduling, the scheduler has to know the state each thread has. This can be solved with signal handling or by writing the current thread state into main memory. In this work, we implement the second approach to avoid having another thread that only schedules tasks to other threads.

Thread safe access to memory can be achieved by different approaches. Firstly, a mutex thread can manage the access for multiple other threads. Secondly, atomic operations ensure the safe access from different threads to the same memory. We use both approaches for different parts of the thread pool. Since using a mutex thread can decrease access performance, we decided to execute atomic operations on memory containing the thread state informations. We use a mutex thread to synchronize the enqueueing of tasks into the priority queue.

\section{Implementation}
In this section, we describe in detail the design and implementation of our thread pool. We show how tasks are enqueued regarding their priority and how the waiting for tasks is implemented. Additionally, we state our thread metrics and show how they are integrated in the thread pool design.

@@

\begin{table*}[htbp]
	\caption{Functions provided by the thread pool to process tasks}
	\begin{center}
		\begin{tabular}{ c c }
			\hline
			\textbf{Function Name}&\textbf{Explanation}\\
			\hline
			thread\_pool\_enqueue\_task & Enqueue task with optional handle \\
			thread\_pool\_enqueue\_tasks & Enqueue tasks with optional handle \\
			thread\_pool\_wait\_for\_task & Waits until the tasks referenced by the handle are completed. \\
			thread\_pool\_enqueue\_tasks\_wait & Enqueue tasks and wait until they are finished. The main thread also participates in task execution\\ 
			thread\_pool\_wait\_for\_all & Wait until all enqueued tasks are finished. The main thread also participates in task execution \\ \hline
		\end{tabular}
		\label{tab1}
	\end{center}
\end{table*}

@@

Each assigned task contains a function pointer to a routine and a pointer to data. We implemented the task queue as a generic priority queue that uses a mutex thread to enqueue tasks threadsafe. Every time a new task is enqueued to the task queue, two scenarios can occur. If the new task has a priority, it is placed in the right slot of the task queue regarding its priority and the subsequent tasks are relocated. If the  new task has no priority, it is appended to the end of the queue. Furthermore, tasks can refer to a handle, which is used by the thread pool to wait for a specific amount of tasks.

\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

In the next step, the tasks will be enqueued into the priority queue of the thread pool. Before enqueueing, the function \emph{thread\_pool\_enqueue\_tasks\_wait} creates a handle and links all tasks to it. After enqueueing the tasks to the priority queue, the waiting threads in the thread pool pop the enqueued tasks from the priority queue and execute the routine of the task. This happens asynchronously to the calling function and therefore is not considered as an own step of the process. The calling thread gets one of the passed tasks and processes it. After finishing the task, the calling thread waits until all other tasks referring to the created handle are finished and then returns to the calling function. In the next section, we show the waiting for tasks with handles in detail.

@@

To solve this problem, we implement the waiting for a set of tasks using a slotmap. In Figure 3, we show the concept of our slotmap. During the enqueueing process of tasks, a set of tasks is mapped to a slot in the slotmap. Each slot contains a number of open tasks to be finished and a generation counter. We name the number of open task \emph{openTaskCount} and the generation counter \emph{genCount}. Every time a set of tasks is enqueued, a slot with $openTaskCount = 0$ is selected and the \emph{genCount} increases. Every time a thread finishes a task, the $openTaskCount$ of the referred slot is decreased. As mentioned before, multiple tasks can refer to a single handle. This handle also refers to a slot and has additionally assigned a generation value. The waiting algorithm loops over the following conditions before returning to the calling function: It checks if the number of open tasks is zero or if the generation counter of this slot has changed. If one of these conditions is true, the waiting algorithm returns to the calling function.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section 2, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates. \\
 This approach enables efficient dynamic and lazy resizing of the thread pool. When the number of threads is reduced, the now redundant threads are marked to finish after completing their current task. If later on the number is increased again, still running but marked threads can be enabled again ensuring that few threads have to be created and that the number of running threads will not temporarily exceed the requested size. \\
 After presenting the enqueueing, processing, and waiting for tasks, we show the second component of our thread pool system, the performance monitoring.

\subsection{Performance monitoring}
The performance monitoring component of the thread pool system is responsible for collecting data about threads, tasks, and the thread pool instance. We will use the information collected in this component for analysis and performance optimization in later stages. As mentioned before, the performance monitoring is an optional component, since evaluating the following statistics can reduce the execution performance of our thread pool. We divide the collected statistics into the sections \emph{task statistics}, \emph{thread statistics}, and \emph{thread pool statistics}.

@@

How long a task waits in the queue before being executed is a good metric to adjust task priority or the thread pool size. For example, if an enqueued task with priority 0 still has a large difference between \emph{enqueue\_time} and \emph{execution\_time}, the thread pool size may be too small. Otherwise, if a large number of tasks with low priority have small differences between \emph{enqueue\_time} and \emph{execution\_time}, this may indicate that the thread pool size is too large. A better indicator of not optimized thread pool sizes are statistics per thread.

@@

The \emph{busy\_time} of a thread strongly depends on the executed tasks and therefore is a metric that gives less information about the thread pool itself. In contrast, the \emph{waiting\_time} is a good indicator to reveal less used threads in the thread pool to determine an optimal amount of threads for the current use case. Furthermore, the \emph{task\_count} of thread compared to the number of tasks processed by the thread pool indicates the degree of usage of the thread. To give further possibilities for optimizing the thread pool, we collect statistics for each thread pool instance.
